---
title: "Starting with data"
author: "Data Carpentry contributors; Kristin Briney; Donna Wrublewski"
minutes: 45
output: html_document
---

------------

> ### Learning Objectives
>
> * Load external data from a .csv file into a data frame.
> * Describe what a data frame is.
> * Summarize the contents of a data frame.
> * Use indexing to subset specific portions of data frames. 
> * Describe what a factor is.
> * Convert between strings and factors.
> * Reorder and rename factors.
> * Change how character strings are handled in a data frame.
> * Format dates.

------------

## Presentation of the Survey Data

We are studying the species repartition and weight of animals caught in plots in our
study area. The dataset is stored as a comma separated value (CSV) file.
Each row holds information for a single animal, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxon            | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

### Downloading Files

Since we already downloaded the data file that we will be using, we don't need 
to follow these directions at this time. However, in the future, you may need to 
access data from an internal server or somewhere on the internet. For example, 
if there is data being periodically written to a server, this command could be 
used in a script to automatically download the most recent version of a data file.

The R function `download.file()` would be used to download the CSV file that 
contains the survey data from Figshare. Inside the download.file command, the 
first entry is a character string with the source URL 
("https://ndownloader.figshare.com/files/2292169"). This source URL downloads a 
CSV file from Figshare. The text after the comma ("data/portal_data_joined.csv") 
is the destination of the file on your local machine. You'll need to have a folder
on your machine called "data" where you'll download the file. So this command 
downloads a file from figshare, names it "portal_data_joined.csv," and adds it to 
a preexisting folder named "data."

Do not run this code chuck - for example only!

```{r, eval=FALSE, purl=TRUE}
# download.file(url="https://ndownloader.figshare.com/files/2292169",
#              destfile = "data/portal_data_joined.csv")
```

### Reading the data into R

The file has now been downloaded to the destination you specified, but R has not yet loaded the data from the file into memory. To do this, we can use the `read_csv()` function from the `tidyverse` package.

Packages in R are basically sets of additional functions that let you do more stuff. The functions we’ve been using so far, like `round()`, `sqrt()`, or `c()` come built into R. Packages give you access to additional functions beyond base R. A similar function to `read_csv()` from the tidyverse package is `read.csv()` from base R. We don’t have time to cover their differences but notice that the exact spelling determines which function is used. Before you use a package for the first time you need to install it on your machine, and then you should import it in every subsequent R session when you need it. We have already done this on Day 1 with the scripts that we ran.

Now, we will use `read_csv()` to load into memory the content of the CSV file as 
an object of class `data.frame` (you CAN run this code chunk):

```{r, eval=TRUE,  purl=FALSE}
surveys <- read_csv("data/portal_data_joined.csv")
```
When you execute `read_csv` on a data file, it looks through the first 1000 rows of each column and guesses its data type. For example, in this dataset, `read_csv()` reads `weight` as `col_double` (a numeric data type), and species as `col_character`. You have the option to specify the data type for a column manually by using the `col_types` argument in `read_csv`.

We can see the contents of the first few lines of the data by typing its name at the command line: `surveys`. By default, this will show you as many rows and columns of the data as fit on your screen. If you wanted the first 50 rows, you could type `print(surveys, n = 50)`.

We can also extract the first few lines of this data using the function `head()`:

```{r, results='show', purl=FALSE}
head(surveys)
## Try also
## View(surveys)
```

> ### Note
>
> `read_csv` assumes that fields are delineated by commas, 
> however, in several countries, the comma is used as a decimal > separator and the semicolon (;) is used as a field 
> delineator. If you want to read in this type of files in R,
> you can use the `read_csv2` function. It behaves exactly like
> `read_csv` but uses different parameters for the decimal and 
> the field separators. There is also the `read_tsv()` for tab separated data files and `read_delim()` for less common formats. Check out the help for `read_csv()` by typing `?read_csv` to learn more. 

> In addition to the above versions of the csv format, you should develop the habits of looking at and recording some parameters of your csv files. For instance, the character encoding, control characters used for line ending, date format (if the date is not split into three variables), and the presence of unexpected [newlines](https://en.wikipedia.org/wiki/Newline) are important characteristics of your data files. Those parameters will ease up the import step of your data in R.

## What are data frames?

When we loaded the data into R, it got stored as an object of class `tibble`, which is a special kind of data frame (the difference is not important for our purposes, but you can learn more about tibbles [here](https://tibble.tidyverse.org/)). Data frames are the de facto data structure for most tabular data, and what we use for statistics and plotting. Data frames can be created by hand, but most commonly they are generated by functions like `read_csv()`; in other words, when importing spreadsheets from your hard drive or the web.

A data frame is the representation of data in the format of a table where the
columns are vectors that all have the same length. Because columns are
vectors, each column must contain a single type of data (e.g., characters, 
integers, factors). For example, here is a figure depicting a data frame 
comprising a numeric, a character, and a logical vector.

![](./img/data-frame.svg)

We can see this when inspecting the <b>str</b>ucture of a data frame
with the function `str()`:

```{r, purl=FALSE}
str(surveys)
```

## Inspecting data frames

We already saw how the functions `head()` and `str()` can be 
useful to check the content and the structure of a data frame. 
Here is a non-exhaustive list of functions to get a sense of 
the content/structure of the data. Let's try them out!

* Size:
    * `dim(surveys)` - returns a vector with the number of rows in the first element, and the number of columns as the second element (the __dim__ensions of
          the object)
    * `nrow(surveys)` - returns the number of rows
    * `ncol(surveys)` - returns the number of columns

* Content:
    * `head(surveys)` - shows the first 6 rows
    * `tail(surveys)` - shows the last 6 rows

* Names:
    * `names(surveys)` - returns the column names (synonym of `colnames()` for `data.frame`
	   objects)
    * `rownames(surveys)` - returns the row names

* Summary:
    * `str(surveys)` - structure of the object and information about the class, length and
	   content of  each column
    * `summary(surveys)` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.

> ### Challenge 3.1

## Indexing and subsetting data frames

Our survey data frame has rows and columns (it has 2 
dimensions). If we want to extract some specific data from it, 
we need to specify the "coordinates" we want from it. Row 
numbers come first, followed by column numbers. However, note
that different ways of specifying these coordinates lead to 
results with different classes.

```{r, purl=FALSE}
# We can extract specific values by specifying row and column 
# indices in the format:
# data_frame[row_index, column_index]
# For instance, to extract the first row and column from 
# the `surveys` data frame:
surveys[1, 1]
 
# First row in the 6th column:
surveys[1, 6]

# We can also use shortcuts to select a number of rows or 
# columns at once.
# To select all columns, leave the column index blank.
# For instance, to select all columns for the first row:
surveys[1,]

# The same shortcut works for rows --
# To select the first column across all rows:
surveys[,1]    

# An even shorter way to select first column across all rows:
surveys[1]

# To select multiple rows or columns, use vectors!
# To select the first three rows of the 5th and 6th column
surveys[c(1, 2, 3), c(5, 6)] 

# We can use the `:` operator to create those vectors for us:
surveys[1:3, 5:6] 

# This is equivalent to head_surveys <- head(surveys)
head_surveys <- surveys[1:6, ] 

# As we've seen, when working with tibbles subsetting with single 
# square brackets ("[ ]") always returns a data frame.
# If you want a vector, use double square brackets ("[[ ]]")

# For instance, to get the first column as a vector:
surveys[[1]]

# To get the first value in our data frame:
surveys[[1, 1]]
```

`:` is a special function that creates numeric vectors of integers in increasing or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain indices of a data frame using the "`-`" sign:

```{r, purl=FALSE}
surveys[, -1]          # The whole data frame, except the first column
surveys[-c(7:34786), ] # Equivalent to head(surveys)
```

Data frames can be subset by calling indices (as shown previously), but also by calling their column names directly:

```{r, eval = FALSE, purl=FALSE}
surveys["species_id"]       # Result is a data frame
surveys[, "species_id"]     # Result is a data frame
surveys[["species_id"]]     # Result is a vector

# We can also use the $ operator with column names instead of double 
# brackets. This returns a vector:
surveys$species_id          # Result is a vector
```

In RStudio, you can use the autocompletion feature to get the full and correct names of the columns.

> ### Challenge 3.2

## Factors

When we did `str(surveys)` we saw that several of the columns consist of integers. The columns `genus`, `species`, `sex`, `plot_type`, … however, are of the class `character`. Arguably, these columns contain categorical data, that is, they can only take on a limited number of values.

R has a special class for working with categorical data, called `factor`. Factors are very useful and actually contribute to making R particularly well suited to working with data. So we are going to spend a little time introducing them.

Once created, factors can only contain a pre-defined set of values, known as *levels*. Factors are stored as integers associated with labels and they can be ordered or unordered. While factors look (and often behave) like character vectors, they are actually treated as integer vectors by R. So you need to be very careful when treating them as strings.

When importing a data frame with `read_csv()`, the columns that contain text are not automatically coerced (=converted) into the `factor` data type, but once we have loaded the data we can do the conversion using the `factor()` function:

```{r, purl=TRUE}
surveys$sex <- factor(surveys$sex)
```

We can see that the conversion has worked by using the summary() function again. This produces a table with the counts for each factor level:

```{r, purl=FALSE}
summary(surveys$sex)
```

By default, R always sorts levels in alphabetical order. For instance, if you have a factor with 2 levels:

```{r, purl=FALSE}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign 1 to the level "female" and 2 to the level "male" (because f comes before m, even though the first element in this vector is "male"). You can see this by using the function `levels()` and you can find the number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you 
might want to specify the order because it is meaningful (e.g., "low", 
"medium", "high"), it improves your visualization, or it is required by 
a particular type of analysis. Here, one way to reorder our levels in 
the `sex` vector would be:

```{r, results=TRUE, purl=FALSE}
sex # current order
sex <- factor(sex, levels = c("male", "female"))
sex # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2, 3), but 
are more informative than integers because factors are self describing: 
`"female"`, `"male"` is more descriptive than `1`, `2`. Which one is 
"male"?  You wouldn't be able to tell just from the integer data. 
Factors, on the other hand, have this information built in. It is 
particularly helpful when there are many levels (like the species names 
in our example dataset).

> # Challenge 3.3

### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(sex)
```

In some cases, you may have to convert factors where the levels appear as numbers (such as concentration levels or years) to a numeric vector. For instance, in one part of your analysis the years might need to be encoded as factors (e.g., comparing average weights across years) but in another part of your analysis they may need to be stored as numeric values (e.g., doing math operations on the years). This conversion from factor to numeric is a little trickier. The `as.numeric()` function returns the index values of the factor,
not its levels, so it will result in an entirely new (and unwanted in this case)
set of numbers. One method to avoid this is to convert factors to characters,
and then to numbers.

Another method is to use the `levels()` function. Compare:

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)               # Wrong! And there is no warning...
as.numeric(as.character(year_fct)) # Works...
as.numeric(levels(year_fct))[year_fct]    # The recommended way.
```

Notice that in the `levels()` approach, three important steps occur:

* We obtain all the factor levels using `levels(year_fct)`
* We convert these levels to numeric values using `as.numeric(levels(year_fct))`
* We then access these numeric values using the underlying integers of the
  vector `year_fct` inside the square brackets

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor
level. Let's look at the number of males and females captured over the course of
the experiment:

```{r, purl=TRUE}
## bar plot of the number of females and males captured during the experiment:
plot(surveys$sex)
```

However, as we saw when we used `summary(surveys$sex)`, there are about 1700 individuals for which the sex information hasn’t been recorded. To show them in the plot, we can turn the missing values into a factor level with the `addNA()` function. We will also have to give the new factor level a label. We are going to work with a copy of the `sex` column, so we’re not modifying the working copy of the data frame:

```{r, results=TRUE, purl=FALSE}
sex <- surveys$sex
levels(sex)
```

```{r, results=TRUE, purl=FALSE}
sex <- addNA(sex)
levels(sex)
```

```{r, results=TRUE, purl=FALSE}
head(sex)
```

```{r, results=TRUE, purl=FALSE}
levels(sex)[3] <- "undetermined"
levels(sex)
```

```{r, results=TRUE, purl=FALSE}
head(sex)
```

Now we can plot the data again, using `plot(sex)`.

```{r, purl=TRUE}
plot(sex)
```

The automatic conversion of data type is sometimes a blessing, sometimes an
annoyance. Be aware that it exists, learn the rules, and double check that data
you import in R are of the correct type within your data frame. If not, use it
to your advantage to detect mistakes that might have been introduced during data
entry (for instance, a letter in a column that should only contain numbers).

Learn more in this [RStudio tutorial](https://support.rstudio.com/hc/en-us/articles/218611977-Importing-Data-with-RStudio)

