---
title: "Formatting Dates"
author: "Kristin Briney; Donna Wrublewski"
output:
  html_document:
    toc: yes
    df_print: paged
  pdf_document: default
---

## Formatting Dates

One of the most common issues that new (and experienced!) R users have is
converting date and time information into a variable that is appropriate and
usable during analyses. As a reminder from earlier in this lesson, the best
practice for dealing with date data is to ensure that each component of your
date is stored as a separate variable. Using `str()`, We can confirm that our
data frame has a separate column for day, month, and year, and that each contains
integer values.

```{r, eval=FALSE, purl=FALSE}
str(surveys)
```

We are going to use the `ymd()` function from the package **`lubridate`** (which 
belongs to the **`tidyverse`**; learn more [here](https://www.tidyverse.org/)). 
**`lubridate`** gets installed as part as the **`tidyverse`** installation. When 
you load  the **`tidyverse`** (`library(tidyverse)`), the core packages (the 
packages used in most data analyses) get loaded. **`lubridate`** however does not 
belong to the core tidyverse, so you have to load it explicitly with 
`library(lubridate)`

Start by loading the required package:

```{r load-package, message=FALSE, purl=FALSE}
library(lubridate)
```

`ymd()` takes a vector representing year, month, and day, and converts it to a
`Date` vector. `Date` is a class of data recognized by R as being a date and can
be manipulated as such. The argument that the function requires is flexible,
but, as a best practice, is a character vector formatted as "YYYY-MM-DD".

Let's create a date object and inspect the structure:

```{r, purl=FALSE}
my_date <- ymd("2015-01-01")
str(my_date)
```

Now let's paste the year, month, and day separately - we get the same result:

```{r, purl=FALSE}
# sep indicates the character to use to separate each component
my_date <- ymd(paste("2015", "1", "1", sep = "-")) 
str(my_date)
```

Now we apply this function to the surveys dataset. Create a character vector from the `year`, `month`, and `day` columns of
`surveys` using `paste()`:

```{r, purl=FALSE}
paste(surveys$year, surveys$month, surveys$day, sep = "-")
```

This character vector can be used as the argument for `ymd()`:

```{r, purl=FALSE}
ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
```

The resulting `Date` vector can be added to `surveys` as a new column called `date`:

```{r, purl=FALSE}
surveys$date <- ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
str(surveys) # notice the new column, with 'date' as the class
```

Let's make sure everything worked correctly. One way to inspect the new column is to use `summary()`:

```{r, results=TRUE, purl=FALSE}
summary(surveys$date)
```

Something went wrong: some dates have missing values. Let's investigate where they 
are coming from.

We can use the functions we saw previously to deal with missing data to identify
the rows in our data frame that are failing. If we combine them with what we 
learned about subsetting data frames earlier, we can extract the columns "year, 
"month", "day" from the records that have `NA` in our new column `date`. We will 
also use `head()` so we don't clutter the output:

```{r, results=TRUE, purl=FALSE}
missing_dates <- surveys[is.na(surveys$date), c("year", "month", "day")]

head(missing_dates)
```

Why did these dates fail to parse? If you had to use these data for your
analyses, how would you deal with this situation?