---
title: "03-starting-with-data-challenges"
author: "Donna T. Wrublewski"
date: "11/4/2021"
output: html_document
---

# Challenge 3.1

Based on the output of `str(surveys)`, can you answer the following questions?

* What is the class of the object `surveys`?
* How many rows and how many columns are in this object?

## Answer

* class: data frame
* how many rows: 34786,  how many columns: 13

# Challenge 3.2

1. Create a `data.frame` (name it `surveys_200`) containing only the data in row 200 of the `surveys` dataset.

2. Notice how `nrow()` gave you the number of rows in a `data.frame`?
* Use that number to pull out just that last row in the data frame.
* Compare that with what you see as the last row using `tail()` to make sure it's meeting expectations.
* Pull out that last row using `nrow()` instead of the row number.
* Create a new data frame (`surveys_last`) from that last row.

3. Use `nrow()` to extract the row that is in the middle of the data frame. Store the content of this row in an object named `surveys_middle`.

4. Combine `nrow()` with the `-` notation above to reproduce the behavior of `head(surveys)`, keeping just the first through 6th rows of the surveys dataset.

## Answer
1. surveys_200 <- surveys[200, ]
2. Saving `n_rows` to improve readability and reduce duplication:
n_rows <- nrow(surveys)
surveys_last <- surveys[n_rows, ]
3. surveys_middle <- surveys[n_rows / 2, ]
4. surveys_head <- surveys[-(7:n_rows), ]

# Challenge 3.3

1. Change the columns `taxa` and `genus` in the `surveys` data frame into a factor.
2. Using the functions you learned before, can you find out…
* How many rabbits were observed?
* How many different genera are in the genus column?

## Answer
surveys$taxa <- factor(surveys$taxa)
surveys$genus <- factor(surveys$genus)
summary(surveys)
nlevels(surveys$genus)

* how many genera: There are 26 unique genera in the `genus` column.
* how many rabbts: There are 75 rabbits in the `taxa` column.

# Challenge 3.4

1. We have seen how data frames are created when using `read.csv()`, but they can also be created by hand with the `data.frame()` function. There are a few mistakes in this hand-crafted `data.frame`. Can you spot and fix them? Don't hesitate to experiment!

>
```{r}
animal_data <- data.frame(
               animal = c(dog, cat, sea cucumber, sea urchin),
               feel = c("furry", "squishy", "spiny"),
               weight = c(45, 8 1.1, 0.8)
               )
```

2. Can you predict the class for each of the columns in the following example? Check your guesses using `str(country_climate)`:
* Are they what you expected?  Why? Why not?
* What would you need to change to ensure that each column had the accurate data type?

```{r}
     country_climate <- data.frame(
            country = c("Canada", "Panama", "South Africa", "Australia"),
            climate = c("cold", "hot", "temperate", "hot/temperate"),
            temperature = c(10, 30, 18, "15"),
            northern_hemisphere = c(TRUE, TRUE, FALSE, "FALSE"),
            has_kangaroo = c(FALSE, FALSE, FALSE, 1)
          )
```

## Answer
1. 
* missing quotations around the names of the animals
* missing one entry in the feel column (probably for one of the furry animals)
* missing one comma in the weight column

2.
* country, climate, temperature, and northern_hemisphere are characters; has_kangaroo is numeric
* using factor() one could replace character columns with factors columns
* removing the quotes in temperature and northern_hemisphere and replacing 1 by TRUE in the has_kangaroo column would give what was probably intended

